dispRity demo

### 2015-10-01 doc/dispRity.demo.tex
Updated demo vignette

\dispRity is a package for calculating disparity in \R.
What is disparity? Well that's a more complex question...
To keep it short, this package allows to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values.
% More explanations to put here?

\subsection{Installation}
You can install this package pretty easily if you use the latest version of \R and \texttt{devtools}.
Just copy past the following
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{install.packages}\hlstd{(}\hlstr{"devtools"}\hlstd{)}
\hlkwd{library}\hlstd{(devtools)}
\hlkwd{install_github}\hlstd{(}\hlstr{"TGuillerme/dispRity"}\hlstd{,} \hlkwc{ref}\hlstd{=}\hlstr{"release"}\hlstd{)}
\hlkwd{library}\hlstd{(dispRity)}
\end{alltt}
\end{kframe}
\end{knitrout}

Note that we use here the \texttt{release} branch which is version 0.1.0.
If you want the piping-hot (and full of bugs) version, change the option \texttt{ref="release"} to \texttt{ref="master"}.

This package depends heavily on the \texttt{ape} package as well as the nice function \texttt{timeSliceTree::paleotree}.
However, I recommend to also install the packages \texttt{geomorph} or \texttt{Claddis} if you want to try this example on your own morphometric or cladistic data.

\subsection{Which data?}
In this demo we are going to use the data from Beck and Lee 2014 \cite{beckancient2014}:

@@

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Loading required package: paleotree}}\begin{alltt}
\hlcom{## An ordinated matrix with the tips only}
\hlkwd{data}\hlstd{(BeckLee_mat50)}
\hlkwd{dim}\hlstd{(BeckLee_mat50)}
\end{alltt}
\begin{verbatim}
## [1] 50 48
\end{verbatim}
\begin{alltt}
\hlkwd{head}\hlstd{(BeckLee_mat50[,}\hlnum{1}\hlopt{:}\hlnum{5}\hlstd{])}
\end{alltt}
\begin{verbatim}
##                    [,1]          [,2]        [,3]       [,4]      [,5]
## Cimolestes   -0.5319679  0.1117759259  0.09865194 -0.1933148 0.2035833
## Maelestes    -0.4087147  0.0139690317  0.26268300  0.2297096 0.1310953
## Batodon      -0.6923194  0.3308625215 -0.10175223 -0.1899656 0.1003108
## Bulaklestes  -0.6802291 -0.0134872777  0.11018009 -0.4103588 0.4326298
## Daulestes    -0.7386111  0.0009001369  0.12006449 -0.4978191 0.4741342
## Uchkudukodon -0.5105254 -0.2420633915  0.44170317 -0.1172972 0.3602273
\end{verbatim}
\begin{alltt}
\hlcom{## An ordinated matrix with the tips and the nodes (named nX)}
\hlkwd{data}\hlstd{(BeckLee_mat99)}
\hlkwd{dim}\hlstd{(BeckLee_mat99)}
\end{alltt}
\begin{verbatim}
## [1] 99 97
\end{verbatim}
\begin{alltt}
\hlkwd{head}\hlstd{(BeckLee_mat99[,}\hlnum{1}\hlopt{:}\hlnum{5}\hlstd{],} \hlnum{2}\hlstd{)}
\end{alltt}
\begin{verbatim}
##                  [,1]       [,2]       [,3]       [,4]        [,5]
## Cimolestes -0.6082437 -0.0323683 0.08458885 -0.4338448 -0.30536875
## Maelestes  -0.5730206 -0.2840361 0.01308847 -0.1258848  0.06123611
\end{verbatim}
\begin{alltt}
\hlkwd{tail}\hlstd{(BeckLee_mat99[,}\hlnum{1}\hlopt{:}\hlnum{5}\hlstd{],} \hlnum{2}\hlstd{)}
\end{alltt}
\begin{verbatim}
##            [,1]      [,2]       [,3]       [,4]       [,5]
## n48 -0.05529018 0.4799330 0.04118477 0.04944912 -0.3558830
## n49 -0.13067785 0.4478168 0.11956268 0.13800340 -0.3222785
\end{verbatim}
\begin{alltt}
\hlcom{## A list of first and last occurence data for some fossils}
\hlkwd{data}\hlstd{(BeckLee_ages)}
\hlkwd{head}\hlstd{(BeckLee_ages)}
\end{alltt}
\begin{verbatim}
##             FAD  LAD
## Adapis     37.2 36.8
## Asioryctes 83.6 72.1
## Leptictis  33.9 33.3
## Miacis     49.0 46.7
## Mimotona   61.6 59.2
## Notharctus 50.2 47.0
\end{verbatim}
\begin{alltt}
\hlcom{## And a phylogeny:}
\hlkwd{data}\hlstd{(BeckLee_tree)}
\hlkwd{plot}\hlstd{(BeckLee_tree) ;} \hlkwd{nodelabels}\hlstd{(}\hlkwc{cex}\hlstd{=}\hlnum{0.8}\hlstd{)}
\end{alltt}
\end{kframe}

@@

However, I greatly encourage you to use your own data (after all, that's the point of an \R package).
You will simply need a ordinated distance matrix (make sure that the number of column is at maximum the number of rows minus one) and a phylogenetic tree (with the tips matching the ordinated matrix).
Just replace all the \texttt{BeckLee} mentions in this demo by your very own data!

\section{A super quick go through}
This package allows the following:
\begin{enumerate}
    \item Splitting the data
    \item Bootstrapping the data
    \item Calculating disparity
    \item Summarising the results
\end{enumerate}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## Splitting the data}
\hlstd{sliced_data} \hlkwb{<-} \hlkwd{time.series}\hlstd{(BeckLee_mat99, BeckLee_tree,} \hlkwc{method} \hlstd{=} \hlstr{"continuous"}\hlstd{,}
    \hlkwc{model} \hlstd{=} \hlstr{"acctran"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlnum{5}\hlstd{,} \hlkwc{FADLAD} \hlstd{= BeckLee_ages)}
\end{alltt}

@@

\hlcom{## Summarising the results}
\hlkwd{summary}\hlstd{(sum_of_ranges)}
\end{alltt}
\begin{verbatim}
##      series  n  mean  2.5%   25%   75% 97.5%
## 1 133.51104  3 17.81  0.00 15.51 24.35 24.35
## 2 100.13328 14 52.05 45.80 49.90 54.50 57.67
## 3  66.75552 23 64.80 59.23 63.20 66.57 69.25
## 4  33.37776 11 42.95 37.94 41.40 44.77 47.43
## 5         0 10 40.94 35.13 39.22 42.69 46.13
\end{verbatim}
\begin{alltt}
\hlkwd{plot}\hlstd{(sum_of_ranges,} \hlkwc{type} \hlstd{=} \hlstr{"continuous"}\hlstd{)}
\end{alltt}
\end{kframe}

@@

Et voil\`{a}...

\section{In more details}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{str}\hlstd{(sum_of_ranges)}
\end{alltt}
\end{kframe}
\end{knitrout}

Therefore this package proposes a new class of object called \dispRity (how unexpected!) that allow to use a \texttt{S3 print.dispRity} method.
In other words, when you are creating objects using the \dispRity functions, you can quickly and easily have a look at them by just printing it.
For example:

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## Which class is the sum_of_ranges object?}
\hlkwd{class}\hlstd{(sum_of_ranges)}
\end{alltt}
\begin{verbatim}
## [1] "dispRity"
\end{verbatim}
\begin{alltt}
\hlcom{## We can summarise it using the S3 method print.dispRity}
\hlstd{sum_of_ranges}
\end{alltt}
\begin{verbatim}
## Disparity measurements across 5 series for 99 taxa. 
## Series:
## 133.51104, 100.13328, 66.75552, 33.37776, 0.
## Disparity calculated as: sum ranges for 97 dimensions.
## Data was split using continuous method.
## Data was bootstrapped 100 times, using the full bootstrap method.
\end{verbatim}
\end{kframe}
\end{knitrout}

Note that each object output from different functions will show summarise slightly different informations.

\subsection{Splitting the data}
One of the first functionality of this package is to facilitate splitting the ordinated matrix in subsamples.
The original ordinated matrix represents the total any-o-space.

\subsubsection{Custom splitting}
\texttt{cust.series} is a fairly straightforward function and allows to split the data according to a factor determined by the user.
For example, here we can split the matrix based on phylogeny: let's classify in two different groups the crown and stem mammals present in the tree.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## We want to separate the species around the node 71 (see phylogeny above).}
\hlcom{## All the descendant are crown and all the ancestors are stem}
\hlstd{crown} \hlkwb{<-} \hlkwd{extract.clade}\hlstd{(BeckLee_tree,} \hlkwc{node} \hlstd{=} \hlnum{71}\hlstd{)}\hlopt{$}\hlstd{tip.label}
\hlstd{stem} \hlkwb{<-} \hlkwd{drop.tip}\hlstd{(BeckLee_tree,} \hlkwc{tip} \hlstd{= crown)}\hlopt{$}\hlstd{tip.label}

\hlcom{## We then have to feed this information in a data.frame with one column}
\hlstd{factors} \hlkwb{<-} \hlkwd{as.data.frame}\hlstd{(}
    \hlkwd{matrix}\hlstd{(}\hlkwc{data} \hlstd{=} \hlkwd{c}\hlstd{(}\hlkwd{rep}\hlstd{(}\hlstr{"crown"}\hlstd{,} \hlkwd{length}\hlstd{(crown)),} \hlkwd{rep}\hlstd{(}\hlstr{"stem"}\hlstd{,} \hlkwd{length}\hlstd{(stem))),}
    \hlkwc{ncol} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{dimnames} \hlstd{=} \hlkwd{list}\hlstd{(}\hlkwd{c}\hlstd{(crown, stem))))}

\hlcom{## Then we can use the customised series function}
\hlstd{crown_stem} \hlkwb{<-} \hlkwd{cust.series}\hlstd{(BeckLee_mat50, factors)}

\hlcom{## This created a dispRity object containing two series: crown and stem}
\hlkwd{class}\hlstd{(crown_stem)}
\end{alltt}
\begin{verbatim}
## [1] "dispRity"
\end{verbatim}
\begin{alltt}
\hlstd{crown_stem}
\end{alltt}
\begin{verbatim}
## 2 custom series for 50 taxa. 
## Series:
## crown, stem.
\end{verbatim}
\begin{alltt}
\hlcom{## This object contains three elements}
\hlkwd{names}\hlstd{(crown_stem)}
\end{alltt}
\begin{verbatim}
## [1] "data"   "taxa"   "series"
\end{verbatim}
\begin{alltt}
\hlcom{## With "data" being the list sub-matrices}
\hlkwd{str}\hlstd{(crown_stem}\hlopt{$}\hlstd{data)}
\end{alltt}
\begin{verbatim}
## List of 2
##  $ crown: num [1:30, 1:48] 0.3079 0.6531 0.5089 -0.1652 -0.0419 ...
##   ..- attr(*, "dimnames")=List of 2
##   .. ..$ : chr [1:30] "Dasypodidae" "Bradypus" "Myrmecophagidae" "Todralestes" ...
##   .. ..$ : NULL
##  $ stem : num [1:20, 1:48] -0.739 -0.68 -0.511 -0.477 -0.473 ...
##   ..- attr(*, "dimnames")=List of 2
##   .. ..$ : chr [1:20] "Daulestes" "Bulaklestes" "Uchkudukodon" "Kennalestes" ...
##   .. ..$ : NULL
\end{verbatim}
\begin{alltt}
\hlcom{## "taxa" being the list of taxa in the original ordinated matrix}
\hlkwd{str}\hlstd{(crown_stem}\hlopt{$}\hlstd{taxa)}
\end{alltt}
\begin{verbatim}
##  chr [1:50] "Cimolestes" "Maelestes" "Batodon" "Bulaklestes" "Daulestes" ...
\end{verbatim}
\begin{alltt}
\hlcom{## "series" containing information on the series type (custom) and names (crown and stem)}
\hlstd{crown_stem}\hlopt{$}\hlstd{series}
\end{alltt}
\begin{verbatim}
## [1] "custom" "crown"  "stem"
\end{verbatim}
\end{kframe}
\end{knitrout}

Of course, using phylogeny as a factor is just an example.
Any type of factors can be used in this example (e.g. body mass categories, habitat, diet, etc...).
Which leads us to another type of category used a lot in disparity-through-time analysis: time!

\subsubsection{Time splitting}
\texttt{time.series} is a function that allows to split the matrix in different ways using time as a category.
This functions needs as input an ordinated matrix and a matching phylogenetic tree.
Two types of time series splitting can be performed with the \dispRity package by using the \texttt{method} option in \texttt{time.series}.
\begin{enumerate}
\item the classical discrete time splitting (or time-binning) using \texttt{method = "discrete"};
\item or a new continuous time splitting (or time-slicing) using \texttt{method = "continuous"}.
\end{enumerate}
For the time-slicing method details, see Guillerme and Cooper \textit{in prep.} \cite{GuillermeSTD}.
Once the method is chosen, the function asks to specify the \texttt{time} which can be given as a vector of \texttt{numeric} values for:
\begin{enumerate}
\item defining the boundaries of the time bins (when \texttt{method = "discrete"});
\item defining the time-slicing places (when using \texttt{method = "continuous"}).
\end{enumerate}
Otherwise, the \texttt{time} argument can be set as a single \texttt{numeric} value for automatically generating a fix number of equidistant time-bins/slices.
Additionally, we can also input a data frame containing the first/last occurrence data (\texttt{FADLAD}) for taxa that span over a longer time than the tips/nodes age.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## Let's generate three time bins containing the data present every 40 Ma}
\hlstd{time_bins} \hlkwb{<-} \hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= BeckLee_mat50,} \hlkwc{tree} \hlstd{= BeckLee_tree,}
    \hlkwc{method} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{120}\hlstd{,} \hlnum{80}\hlstd{,} \hlnum{40}\hlstd{,} \hlnum{0}\hlstd{))}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# No FADLAD table has been provided so every tip is assumed to interval single points in time.}}\begin{alltt}
\hlcom{## Note that the function provides a warning saying that tips where single}
\hlcom{## points in time (no FAD/LAD information). We can fix that by adding the}
\hlcom{## age data for the taxa that have some longer occurrence spans.}
\hlstd{time_bins} \hlkwb{<-} \hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= BeckLee_mat50,} \hlkwc{tree} \hlstd{= BeckLee_tree,}
    \hlkwc{method} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{120}\hlstd{,} \hlnum{80}\hlstd{,} \hlnum{40}\hlstd{,} \hlnum{0}\hlstd{),} \hlkwc{FADLAD} \hlstd{= BeckLee_ages)}
\end{alltt}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= BeckLee_mat50,} \hlkwc{tree} \hlstd{= BeckLee_tree,} \hlkwc{method} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlnum{3}\hlstd{)}
\end{alltt}

>>>>

\dispRity is a package for calculating disparity in \texttt{R}.
To keep it short, this package allows to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) into single values.
These ordinated matrices can be called the morphospace (when using morphometric data) or the cladisto-space (for cladistic data) and will represent the total spread of the forms available in the dataset.
However, keep in mind that if you are using other type of data, say ecological data for example, then the ordinated matrix will represent the eco-space which represents the total spread of ecological traits available in the dataset.
In the rest of this demo, I will sometimes refer to the ordinated matrix as the any-o-space where you can replace \textit{any} by any prefix you want your space to be based on!

\subsection{Installation}
You can install this package easily if you use the latest version of \R and \texttt{devtools}.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{install.packages}\hlstd{(}\hlstr{"devtools"}\hlstd{)}
\hlkwd{library}\hlstd{(devtools)}
\hlkwd{install_github}\hlstd{(}\hlstr{"TGuillerme/dispRity"}\hlstd{,} \hlkwc{ref} \hlstd{=} \hlstr{"release"}\hlstd{)}
\hlkwd{library}\hlstd{(dispRity)}
\end{alltt}
\end{kframe}
\end{knitrout}

Note that we use the \texttt{release} branch here which is version 0.1.0.
If you want the piping-hot (and full of bugs) version, change the option \texttt{ref = "release"} to \texttt{ref = "master"}.
This package depends mainly on the \texttt{ape} package and the \texttt{timeSliceTree::paleotree} function.
Future version will just require the specific function but this version (0.1.0) will install \texttt{paleotree} as a dependency (which is not that bad since it is a super useful package anyway!).
Additionally, I warmly recommend to install the packages \texttt{geomorph} or/and \texttt{Claddis} if you want to try this demo with your very own ordinated morphometric or cladistic data.

\subsection{Which data?}
In this demo we are going to use a subset of the ordinated cladistic data from Beck and Lee 2014 \cite{beckancient2014}:

@@

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Loading required package: paleotree}}\begin{alltt}
\hlcom{## An ordinated matrix with the tips only}
\hlkwd{data}\hlstd{(BeckLee_mat50)}
\hlkwd{dim}\hlstd{(BeckLee_mat50)}
\end{alltt}
\begin{verbatim}
## [1] 50 48
\end{verbatim}
\begin{alltt}
\hlkwd{head}\hlstd{(BeckLee_mat50[,}\hlnum{1}\hlopt{:}\hlnum{5}\hlstd{])}
\end{alltt}
\begin{verbatim}
##                    [,1]          [,2]        [,3]       [,4]      [,5]
## Cimolestes   -0.5319679  0.1117759259  0.09865194 -0.1933148 0.2035833
## Maelestes    -0.4087147  0.0139690317  0.26268300  0.2297096 0.1310953
## Batodon      -0.6923194  0.3308625215 -0.10175223 -0.1899656 0.1003108
## Bulaklestes  -0.6802291 -0.0134872777  0.11018009 -0.4103588 0.4326298
## Daulestes    -0.7386111  0.0009001369  0.12006449 -0.4978191 0.4741342
## Uchkudukodon -0.5105254 -0.2420633915  0.44170317 -0.1172972 0.3602273
\end{verbatim}
\begin{alltt}
\hlcom{## An ordinated matrix with the tips and the nodes (named nX)}
\hlkwd{data}\hlstd{(BeckLee_mat99)}
\hlkwd{dim}\hlstd{(BeckLee_mat99)}
\end{alltt}
\begin{verbatim}
## [1] 99 97
\end{verbatim}
\begin{alltt}
\hlkwd{head}\hlstd{(BeckLee_mat99[,}\hlnum{1}\hlopt{:}\hlnum{5}\hlstd{],} \hlnum{2}\hlstd{)}
\end{alltt}
\begin{verbatim}
##                  [,1]       [,2]       [,3]       [,4]        [,5]
## Cimolestes -0.6082437 -0.0323683 0.08458885 -0.4338448 -0.30536875
## Maelestes  -0.5730206 -0.2840361 0.01308847 -0.1258848  0.06123611
\end{verbatim}
\begin{alltt}
\hlkwd{tail}\hlstd{(BeckLee_mat99[,}\hlnum{1}\hlopt{:}\hlnum{5}\hlstd{],} \hlnum{2}\hlstd{)}
\end{alltt}
\begin{verbatim}
##            [,1]      [,2]       [,3]       [,4]       [,5]
## n48 -0.05529018 0.4799330 0.04118477 0.04944912 -0.3558830
## n49 -0.13067785 0.4478168 0.11956268 0.13800340 -0.3222785
\end{verbatim}
\begin{alltt}
\hlcom{## A list of first and last occurrences data for some fossils}
\hlkwd{data}\hlstd{(BeckLee_ages)}
\hlkwd{head}\hlstd{(BeckLee_ages)}
\end{alltt}
\begin{verbatim}
##             FAD  LAD
## Adapis     37.2 36.8
## Asioryctes 83.6 72.1
## Leptictis  33.9 33.3
## Miacis     49.0 46.7
## Mimotona   61.6 59.2
## Notharctus 50.2 47.0
\end{verbatim}
\begin{alltt}
\hlcom{## And a phylogeny:}
\hlkwd{data}\hlstd{(BeckLee_tree)}
\hlkwd{plot}\hlstd{(BeckLee_tree) ;} \hlkwd{nodelabels}\hlstd{(}\hlkwc{cex}\hlstd{=}\hlnum{0.8}\hlstd{)}
\end{alltt}
\end{kframe}

@@

However, I greatly encourage you to use your own data (after all, that's the point of an \R package).
You will simply need (1) an ordinated distance matrix: make sure that the number of column is at maximum the number of rows minus one.
And secondly (2) a phylogenetic tree with the tips matching the ordinated matrix and a \texttt{\$root.time} element for determining the age of the tree.
You can use the function \texttt{tree.age} provided in this package but not described in this demo for generating the root.time automatically (see \texttt{?tree.age}).
In the following examples, just replace all the \texttt{BeckLee} mentions in this demo by your very own data:

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## My ordinated matrix with tips only}
\hlstd{MyData_mat50} \hlkwb{<-} \hlstd{BeckLee_mat50}

\hlcom{## My ordinated matrix with tips and nodes}
\hlstd{MyData_mat99} \hlkwb{<-} \hlstd{BeckLee_mat99}

\hlcom{## My first and last occurence data}
\hlstd{MyData_ages} \hlkwb{<-} \hlstd{BeckLee_ages}

\hlcom{## My phylogenetic tree}
\hlstd{MyData_tree} \hlkwb{<-} \hlstd{BeckLee_tree}
\end{alltt}
\end{kframe}
\end{knitrout}

\section{A super quick go through}
Without much explanation, here's what this package allows to run:
\begin{enumerate}
    \item Splitting the data
    \item Bootstrapping the data
    \item Calculating disparity
    \item Summarising the results
\end{enumerate}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## Splitting the data}
\hlstd{sliced_data} \hlkwb{<-} \hlkwd{time.series}\hlstd{(MyData_mat99, MyData_tree,} \hlkwc{method} \hlstd{=} \hlstr{"continuous"}\hlstd{,}
    \hlkwc{model} \hlstd{=} \hlstr{"acctran"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlnum{5}\hlstd{,} \hlkwc{FADLAD} \hlstd{= MyData_ages)}
\end{alltt}

@@

\hlcom{## Summarising the results}
\hlkwd{summary}\hlstd{(sum_of_ranges)}
\end{alltt}
\begin{verbatim}
##      series  n  mean  2.5%   25%   75% 97.5%
## 1 133.51104  3 16.67  0.00 15.51 18.67 24.35
## 2 100.13328 14 51.48 45.03 49.29 53.62 56.54
## 3  66.75552 23 64.77 60.09 63.26 66.34 69.58
## 4  33.37776 11 42.41 35.87 40.94 44.40 46.98
## 5         0 10 40.20 34.46 38.38 42.46 44.57
\end{verbatim}
\begin{alltt}
\hlkwd{plot}\hlstd{(sum_of_ranges,} \hlkwc{type} \hlstd{=} \hlstr{"continuous"}\hlstd{)}
\end{alltt}
\end{kframe}

@@

\textit{Et voil\`{a}}, that's pretty much it...

\section{In more details}
In this section we are going to look at each functionality in more details. Of course, at this stage of the development of the package, any suggestion is most welcome.
Please create an issue on the \href{https://github.com/TGuillerme/dispRity}{GitHub page: github.com/TGuillerme/dispRity} or even just drop me an email at \href{mailto:guillert@tcd.ie}{guillert@tcd.ie}.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{str}\hlstd{(sum_of_ranges)}
\hlcom{## That's a 2512 lines output!}
\end{alltt}
\end{kframe}
\end{knitrout}

Therefore this package proposes a new class of object called \dispRity (how unexpected!) that allow to use a S3 \texttt{print.dispRity} method.
In other words, when you are creating objects using the \dispRity functions, you can quickly and easily have a look at them by just calling the object (it will automatically print it via \texttt{print.dispRity}).
For example:

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## Which class is the sum_of_ranges object?}
\hlkwd{class}\hlstd{(sum_of_ranges)}
\end{alltt}
\begin{verbatim}
## [1] "dispRity"
\end{verbatim}
\begin{alltt}
\hlcom{## We can summarise it using the S3 method print.dispRity}
\hlstd{sum_of_ranges}
\end{alltt}
\begin{verbatim}
## Disparity measurements across 5 series for 99 taxa. 
## Series:
## 133.51104, 100.13328, 66.75552, 33.37776, 0.
## Disparity calculated as: sum ranges for 97 dimensions.
## Data was split using continuous method.
## Data was bootstrapped 100 times, using the full bootstrap method.
\end{verbatim}
\begin{alltt}
\hlcom{## Some attentive ape package users will note the similarities with a phylo object...}
\end{alltt}
\end{kframe}
\end{knitrout}

Note that each object output from the different functions detailed below will display slightly different summarised informations (depending on the object content).

\subsection{Splitting the data}
One of the first functionality of this package is to facilitate splitting the ordinated matrix in subsamples.
In fact, in many disparity analysis, one will analysis differences in subsamples of the any-o-space.
For example, the subsamples can be composed of elements at a certain point in time (disparity-through-time analysis), or elements with some traits in common (e.g. habitat, phylogeny, etc.).

\subsubsection{Custom splitting}
\texttt{cust.series} is a function that allows to split the ordinated matrix according to a factor determined by the user.
For example, here we can split the matrix based on phylogeny: let's classify in two different groups the crown and stem mammals present in the tree.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## We want to separate the species around the node 71 (see phylogeny above).}
\hlcom{## All the descendant of these node are crown and all the ancestors are stem}
\hlstd{crown} \hlkwb{<-} \hlkwd{extract.clade}\hlstd{(MyData_tree,} \hlkwc{node} \hlstd{=} \hlnum{71}\hlstd{)}\hlopt{$}\hlstd{tip.label}
\hlstd{stem} \hlkwb{<-} \hlkwd{drop.tip}\hlstd{(MyData_tree,} \hlkwc{tip} \hlstd{= crown)}\hlopt{$}\hlstd{tip.label}

\hlcom{## We then have to feed this information in a data frame with one column}
\hlstd{factors} \hlkwb{<-} \hlkwd{as.data.frame}\hlstd{(}
    \hlkwd{matrix}\hlstd{(}\hlkwc{data} \hlstd{=} \hlkwd{c}\hlstd{(}\hlkwd{rep}\hlstd{(}\hlstr{"crown"}\hlstd{,} \hlkwd{length}\hlstd{(crown)),} \hlkwd{rep}\hlstd{(}\hlstr{"stem"}\hlstd{,} \hlkwd{length}\hlstd{(stem))),}
    \hlkwc{ncol} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{dimnames} \hlstd{=} \hlkwd{list}\hlstd{(}\hlkwd{c}\hlstd{(crown, stem))))}

\hlcom{## We then can use the customised series function}
\hlstd{crown_stem} \hlkwb{<-} \hlkwd{cust.series}\hlstd{(MyData_mat50, factors)}

\hlcom{## This created a dispRity object containing two series: crown and stem}
\hlkwd{class}\hlstd{(crown_stem)}
\end{alltt}
\begin{verbatim}
## [1] "dispRity"
\end{verbatim}
\begin{alltt}
\hlstd{crown_stem}
\end{alltt}
\begin{verbatim}
## 2 custom series for 50 taxa. 
## Series:
## crown, stem.
\end{verbatim}
\begin{alltt}
\hlcom{## This object contains three elements}
\hlkwd{names}\hlstd{(crown_stem)}
\end{alltt}
\begin{verbatim}
## [1] "data"   "taxa"   "series"
\end{verbatim}
\begin{alltt}
\hlcom{## With "data" being the list sub-matrices}
\hlkwd{str}\hlstd{(crown_stem}\hlopt{$}\hlstd{data)}
\end{alltt}
\begin{verbatim}
## List of 2
##  $ crown: num [1:30, 1:48] 0.3079 0.6531 0.5089 -0.1652 -0.0419 ...
##   ..- attr(*, "dimnames")=List of 2
##   .. ..$ : chr [1:30] "Dasypodidae" "Bradypus" "Myrmecophagidae" "Todralestes" ...
##   .. ..$ : NULL
##  $ stem : num [1:20, 1:48] -0.739 -0.68 -0.511 -0.477 -0.473 ...
##   ..- attr(*, "dimnames")=List of 2
##   .. ..$ : chr [1:20] "Daulestes" "Bulaklestes" "Uchkudukodon" "Kennalestes" ...
##   .. ..$ : NULL
\end{verbatim}
\begin{alltt}
\hlcom{## "taxa" being the list of taxa in the original ordinated matrix}
\hlkwd{str}\hlstd{(crown_stem}\hlopt{$}\hlstd{taxa)}
\end{alltt}
\begin{verbatim}
##  chr [1:50] "Cimolestes" "Maelestes" "Batodon" "Bulaklestes" "Daulestes" ...
\end{verbatim}
\begin{alltt}
\hlcom{## and finally, "series" containing information on the series type (custom) and names}
\hlcom{## (crown and stem)}
\hlstd{crown_stem}\hlopt{$}\hlstd{series}
\end{alltt}
\begin{verbatim}
## [1] "custom" "crown"  "stem"
\end{verbatim}
\end{kframe}
\end{knitrout}

Of course, using phylogeny as a factor is just an example.
Any type of factors can be used in this example (e.g. body mass categories, habitat, diet, etc...).

\subsubsection{Time splitting}
\texttt{time.series} is a function that allows to split the matrix in different ways using time as a category.
This functions needs as input an ordinated matrix and a matching phylogenetic tree.
Two types of time series splitting can be performed by using the \texttt{method} option:
\begin{enumerate}
\item the classical discrete time splitting (or time-binning) using \texttt{method = "discrete"};
\item or a new continuous time splitting (or time-slicing) using \texttt{method = "continuous"}.
\end{enumerate}
For the time-slicing method details, see Guillerme and Cooper \textit{in prep.} \cite{GuillermeSTD}.
Once the method is chosen, the function intakes the \texttt{time} argument which can be a vector of \texttt{numeric} values for:
\begin{enumerate}
\item defining the boundaries of the time bins (when \texttt{method = "discrete"});
\item defining the time-slicing places (when using \texttt{method = "continuous"}).
\end{enumerate}
Otherwise, the \texttt{time} argument can be set as a single \texttt{numeric} value for automatically generating a fix number of equidistant time-bins/slices.
Additionally, we can also input a data frame containing the first/last occurrence data (\texttt{FADLAD}) for taxa that span over a longer time than the tips/nodes age.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## Let's generate three time bins containing the data present every 40 Ma}
\hlstd{time_bins} \hlkwb{<-} \hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= MyData_mat50,} \hlkwc{tree} \hlstd{= MyData_tree,}
    \hlkwc{method} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{120}\hlstd{,} \hlnum{80}\hlstd{,} \hlnum{40}\hlstd{,} \hlnum{0}\hlstd{))}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# No FADLAD table has been provided so every tip is assumed to interval single points in time.}}\begin{alltt}
\hlcom{## Note that the function provides a warning saying that tips where single}
\hlcom{## points in time (no FAD/LAD information). We can fix that by adding the}
\hlcom{## age data for the taxa that have some longer occurrence spans.}
\hlstd{time_bins} \hlkwb{<-} \hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= MyData_mat50,} \hlkwc{tree} \hlstd{= MyData_tree,}
    \hlkwc{method} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlkwd{c}\hlstd{(}\hlnum{120}\hlstd{,} \hlnum{80}\hlstd{,} \hlnum{40}\hlstd{,} \hlnum{0}\hlstd{),} \hlkwc{FADLAD} \hlstd{= MyData_ages)}
\end{alltt}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{time.series}\hlstd{(}\hlkwc{data} \hlstd{= MyData_mat50,} \hlkwc{tree} \hlstd{= MyData_tree,} \hlkwc{method} \hlstd{=} \hlstr{"discrete"}\hlstd{,} \hlkwc{time} \hlstd{=} \hlnum{3}\hlstd{)}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# No FADLAD table has been provided so every tip is assumed to interval single points in time.}}\begin{verbatim}
## 3 discrete series for 50 taxa. 
## Series:
## 133.51104-89.00736, 89.00736-44.50368, 44.50368-0.
\end{verbatim}
\end{kframe}
\end{knitrout}
